Examining Programming Students’ Interaction with Generative AI: A Metacognitive Perspective

### 2024-10-01 CSEDU2025/Paper.tex

\title{
  Temporary: An Analysis of Programming Students’ Interaction Prompts with
  Generative AI: A Metacognitive Perspective
}

\author{\authorname{Rodrigo Prestes Machado\sup{1}\orcidAuthor{0000-0003-0428-6387},
Carlos Alario Hoyos\sup{2}\orcidAuthor{0000-0002-3082-0814},
Patricia Callejo Pinardo\sup{2}\orcidAuthor{0000-0001-6124-6213},
Iria Manuela Estévez Ayres\sup{2}\orcidAuthor{0000-0002-1047-5398},
Carlos Delgado Kloos\sup{2}\orcidAuthor{0000-0003-4093-3705}}
\affiliation{\sup{1}Department of Informatics, Federal Institute of Education,
Science and Technology, Porto Alegre, Brazil}
\affiliation{\sup{2}Department of Telematics Engineering, Carlos III University,
Madrid, Spain}
\email{\ rodrigo.prestes@poa.ifrs.edu.br, \{calario,pcallejo, ayres,cdk\}@it.uc3m.es}
}

@@

\abstract{The abstract should summarize the contents of the paper and should
contain at least 70 and at most 200 words. The text must be set to 9-point font
size.}

@@

Recent advancements in Generative Artificial Intelligence (GenAI) have opened up
new possibilities in education. In programming courses, students can utilize
GenAI tools to improve their understanding, receive feedback, and obtain
detailed explanations.

@@

Regardless of teachers' preferences or beliefs, a preliminary surveys performed
by \cite{Dickey24} revealed that at least 54.5\% of students are already using
GenAI for homework. This highlights the need to increase the understanding of
how students interact with these tools and how they can be used to
enhance learning. Furthermore, a review organized by \cite{Lo24} emphasized the
need for extended studies and objective measures to gather more robust
evidence on the use of GenAI tools in education.

@@

\begin{enumerate}
  \item What was the distribution pattern of the classified messages?
  \item Was there a space between student prompts? If so, which category promote the largest spacing between submissions?
  \item Do students' conversations with the bot alternate between practical and theoretical approaches, thereby promoting interleaving?
\end{enumerate}

@@

In the categorization phase, the students' messages were manually classified
into eight distinct categories for analysis. Initially, \cite{Ghimire24}
proposed four categories: Debugging Help, Conceptual Question, Code Snippet,
and Complete Solution. However, the data collected indicated the need for
additional categories, leading to the inclusion of four more: Multiple Choice
(Quiz), Student Corrections, Language Change, and Uncategorized. In this study,
students were free to use the bot as they wished, which contributed to the
emergence of these additional categories. Table \ref{tab:categories} presents
these categories along with their respective descriptions.

@@

A total of 343 students' messages were categorized across 42 conversations, with
an average of 8.17 messages per conversation. This data offers insights into how
users engage with CharlieBot and the types of prompts they submit. The analysis
results are presented in this section, addressing the research questions
outlined in the introduction.

The Figure \ref{fig:graph1} shows the distribution of the messages in the
categories:

@@

According to Figure \ref{fig:graph1} messages classified as
Conceptual Question account for 32.1\% of the total. Often, the sequence of
Conceptual Question begins with a request for a Complete Solution or Code
Snippet, which then evolves into a series of Conceptual Question, as seen in
conversation 2 of Figure \ref{fig:graph2}. In other cases, the conversation
consists entirely of messages classified as Conceptual Question, such as in
conversation 10 of Figure \ref{fig:graph2}. This type of message indicates that
students recognize they have not yet fully mastered a concept and are seeking to
better understand it. In this context, the student realizes that solving the
problem requires understanding of concepts, thus applying an active learning
strategy.

As shown in Figure \ref{fig:graph1}, approximately 30.9\% of the messages were
classified as Debugging Help. These prompts often begin with a request for a
Complete Solution or Code Snippet. Following this, students typically switch from
passive to more active learning strategy, as illustrated by the start
of conversation 12 in Figure \ref{fig:graph2}. Debugging Help messages represent
a form of active learning, where students likely focus on gaining a practical
understanding of the code, identifying errors, and improving their debugging
skills for future tasks.

@@

Student corrections to bot responses accounted for 2.6\% of the
interactions. In most cases, these corrections occurred when the student already
had the correct answer to an exercise and identified an error in the bot’s
response, as illustrated in conversations 4 and 9 in Figure \ref{fig:graph2}.
These corrections highlight that, although the bot provides quick feedback, it
is not always accurate, and its errors can be difficult for students to
identify.

The Figure \ref{fig:graph1} also shows that around 5\% of the prompts written by
students were not categorized. These messages range from greetings to the Bot,
such as a simple "\textit{Hola}," to contextual statements like
"\textit{Estoy repasando orientación a objetos.}"

@@

About the second research question: \textit{Was there a space between student
prompts? If so, which category promotes the largest spacing between
submissions?}

Spacing, a well-established learning strategy \citep{Carvalho20}, was analyzed
by examining the time intervals between users' interactions with CharlieBot.
Only 9\% of the conversations showed message delays exceeding 24 hours,
indicating that students generally do not extend interactions across multiple
days. However, this students' behavior is directly influenced by the
implementation of the initial version of CharlieBot, which was designed to keep
the chat session active only while the student kept the browser tab open and
avoided reloading the page.

On the other hand, about 26.2\% of the conversations, or roughly a quarter,
included pauses longer than 60 minutes, a practice known to help with knowledge
retention. However, this data only shows that students paused their use of the
AI, which could mean they were reflecting on the topic independently or engaging
in unrelated activities, making it uncertain in what context the students were
applying spacing.

There is a noticeable difference in the interaction time with the bot when a
student asks a conceptual question. This was probably due to the fact that, in
this category, the student is more focused on understanding a concept and
engaging in deeper learning, which aligns with the metacognitive goal of
promoting comprehension and long-term retention. The Figure \ref{fig:graph3}
shows the average interaction time per category.

@@

When analyzing the classification of messages in conversations, we identified
the presence of sequences of messages belonging to the same category, that it
could suggest a low occurrence of interleaving. To illustrate the presence of
blocks the Figure \ref{fig:graph2} presents examples of conversations and their
classification. In Conversation 2, the student alternates messages among Code
Snippet, Debugging Help and Conceptual Question, showing an tendency of
interleaving from practical to theoretical perspective. In contrast, in
Conversation 10, the student focused on a single block of Conceptual Question.
Although the focus was on understanding a theoretical concept, this conversation
tended not to promote interleaving, due to the absence of alternation between
different types of tasks.

To better assess the extent of these sequences, we calculated the percentage of
messages that were part of repeated blocks. Considering blocks with at least
three messages, we observed that 56.2\% of the students' prompts were grouped
into some sequence. When analyzing blocks with a minimum of four messages, we
found that 47.5\% of the prompts were part of continuous sequences.

However, if we consider that sequential messages from a certain point indicate
that the student is \textit{blocked} (or focused) on a task, the remaining space
for interleaving is 43.8\%, assuming a block starting from at least three
consecutive messages, and 52.5\% for blocks starting from four consecutive
messages. Thus, the data suggest that interleaving occurs on students'
conversations with CharlieBot but in which situations and with what type of
students profile is a study that needs to be further investigated.

@@

* Although the data indicates some use of interleaving and spacing strategies,
we believe that students can be guided to adopt a more deliberate approach in
utilizing GenAI, enabling them to more effectively harness these tools to
enhance their learning.

>>>>

\title{
  Examining Programming Students’ Interaction with Generative AI: A
  Metacognitive Perspective
}

\author{\authorname{Rodrigo Prestes Machado\sup{1}\orcidAuthor{0000-0003-0428-6387},
Carlos Alario Hoyos\sup{2}\orcidAuthor{0000-0002-3082-0814},
Patricia Callejo Pinardo\sup{2}\orcidAuthor{0000-0001-6124-6213},
Iria Estévez-Ayres\sup{2}\orcidAuthor{0000-0002-1047-5398},
Carlos Delgado Kloos\sup{2}\orcidAuthor{0000-0003-4093-3705}}
\affiliation{\sup{1}Department of Informatics, Instituto Federal de Educação,
Ciência e Tecnologia, Porto Alegre, Brazil}
\affiliation{\sup{2}Department of Telematics Engineering, Universidad Carlos III
de Madrid, Leganés (Madrid), Spain}
\email{\ rodrigo.prestes@poa.ifrs.edu.br, \{calario,pcallejo, ayres,cdk\}@it.uc3m.es}
}

@@

\abstract{This study investigates the interaction patterns of programming
students with a generative AI-powered chatbot (CharlieBot) in the context of a
Java programming course. Specifically, it explores how students utilize the bot,
enhanced with Retrieval-Augmented Generation (RAG), along with two metacognitive
strategies: interleavBaha24Sun24ing and spacing. The methodology analyzed and
classified student conversations, focusing on types of interactions such as
conceptual questions and debugging help. The results indicate that most
interactions are focused on active learning strategies, although the deliberate use
of interleaving and spacing is limited. The study highlights the need for more
guidance on the strategic use of AI tools to improve learning outcomes.}

@@

Recent advances in Generative Artificial Intelligence (GenAI) have opened up
new possibilities in education. In programming courses, students can utilize
GenAI tools to improve their understanding, receive feedback, and obtain
detailed explanations.

@@

Regardless of teachers' preferences or beliefs, preliminary surveys conducted
by \cite{Dickey24} revealed that at least 54.5\% of students are already using
GenAI for homework. This highlights the need to increase the understanding of
how students interact with these tools and how they can be used to
enhance learning. Furthermore, a review organized by \cite{Lo24} emphasized the
need for extended studies and objective measures to gather more robust
evidence on the use of GenAI tools in education.

@@

\begin{enumerate}
  \item What was the distribution pattern of the classified messages?
  \item Was there a space between the student prompts? If so, which category promotes the largest spacing between submissions?
  \item Do students' conversations with the bot alternate between practical and theoretical approaches, thereby promoting interleaving?
\end{enumerate}

@@

In the categorization phase, the students' messages were manually classified
into eight distinct categories for analysis. Initially, \cite{Ghimire24}
proposed four categories: Debugging Help, Conceptual Question, Code Snippet,
and Complete Solution. However, the data collected indicated the need for
additional categories, leading to the inclusion of four more: Multiple Choice
(Quiz), Student Corrections, Language Change, and Uncategorized. In this study,
students were free to use the bot as they wanted, which contributed to the
emergence of these additional categories. Table \ref{tab:categories} presents
these categories along with their respective descriptions.

@@

A total of 343 student messages were categorized in 42 conversations, with
an average of 8.17 messages per conversation. This data offers insights into how
users engage with CharlieBot and the types of prompts they submit. The results of
the analysis are presented in this section, addressing the research questions
outlined in the introduction.

Figure \ref{fig:graph1} shows the distribution of the messages in the
categories:

@@

According to Figure \ref{fig:graph1} messages classified as
Conceptual Question account for 32.1\% of the total. Often, the sequence of
Conceptual Question begins with a request for a Complete Solution or Code
Snippet, which then evolves into a series of Conceptual Question, as seen in
conversation 2 of Figure \ref{fig:graph2}. In other cases, the conversation
consists entirely of messages classified as Conceptual Question, such as in
conversation 10 of Figure \ref{fig:graph2}. This type of message indicates that
students recognize that they have not yet fully mastered a concept and are seeking to
better understand it. In this context, the student realizes that solving the
problem requires understanding of concepts, thus applying an active learning
strategy.

As shown in Figure \ref{fig:graph1}, approximately 30.9\% of the messages were
classified as Debugging Help. These prompts often begin with a request for a
Complete Solution or Code Snippet. Following this, students typically switch from
passive to more active learning strategy, as illustrated by the start
of conversation 12 in Figure \ref{fig:graph2}. Debugging Help messages represent
a form of active learning, in which students likely focus on gaining a practical
understanding of the code, identifying errors, and improving their debugging
skills for future tasks.

@@

Student corrections to bot responses accounted for 2.6\% of the
interactions. In most cases, these corrections occurred when the student already
had the correct answer to an exercise and identified an error in the bot
response, as illustrated in conversations 4 and 9 in Figure \ref{fig:graph2}.
These corrections highlight that, although the bot provides quick feedback, it
is not always accurate and its errors can be difficult for students to
identify.

Figure \ref{fig:graph1} also shows that around 5\% of the prompts written by
students were not categorized. These messages range from greetings to the Bot,
such as a simple "\textit{Hola}," to contextual statements like
"\textit{Estoy repasando orientación a objetos.}"

@@

About the second research question: \textit{Was there a space between the student
prompts? If so, which category promotes the largest spacing between
submissions?}

Spacing, a well-established learning strategy \citep{Carvalho20}, was analyzed
by examining the time intervals between user interactions with CharlieBot.
Approximately 26.2\% of the conversations, or roughly a quarter, included pauses
longer than 60 minutes. However, this data only indicates that students paused
their use of the AI, which could suggest that they were either reflecting on the
topic independently or engaging in unrelated activities. Furthermore, 16.7\%
of the conversations exhibited pauses of 12 hours and 9\% showed delays
exceeding 24 hours between messages, indicating that some degree of spacing
occurred. However, it is important to note that this behavior may have been
influenced by the design of the initial version of CharlieBot, which only kept
the chat session active as long as the student kept the browser tab open and
avoided reloading the page. This adds uncertainty about the precise context in
which students applied spacing.

There is a noticeable difference in the interaction time with the bot when a
student asks a conceptual question. This was probably due to the fact that, in
this category, the student is more focused on understanding a concept and
engaging in deeper learning, which aligns with the metacognitive goal of
promoting comprehension and long-term retention. Figure \ref{fig:graph3}
shows the average interaction time per category.

@@

When analyzing the classification of messages in conversations, we identified
the presence of sequences of messages belonging to the same category, which
could suggest a low occurrence of interleaving. To illustrate the presence of
blocks, Figure \ref{fig:graph2} presents examples of conversations and their
classification. In Conversation 2, the student alternates messages among Code
Snippet, Debugging Help, and Conceptual Question, showing a tendency to
interleave from a practical to theoretical perspective. In contrast, in
Conversation 10, the student focused on a single block of Conceptual Question.
Although the focus was on understanding a theoretical concept, this conversation
tended not to promote interleaving, due to the absence of alternation between
different types of task.

To better assess the extent of these sequences, we calculated the percentage of
messages that were part of repeated blocks. Considering blocks with at least
three messages, we observed that 56.2\% of the student's prompts were grouped
into some sequence. When analyzing blocks with a minimum of four messages, we
found that 47.5\% of the prompts were part of continuous sequences.

However, if we consider that sequential messages from a certain point indicate
that the student is \textit{blocked} (or focused) on a task, the remaining space
for interleaving is 43.8\%, assuming a block starting from at least three
consecutive messages, and 52.5\% for blocks starting from four consecutive
messages. Thus, the data suggest that interleaving occurs on students'
conversations with CharlieBot, but in which situations and with what type of
student profile is a study that needs to be further investigated.

@@

* The results indicate that most interactions are focused on active learning
strategies, although deliberate use of interleaving and spacing is limited. We
believe that students can be guided to adopt a more deliberate approach in
utilizing GenAI, enabling them to more effectively harness these tools to
enhance their learning.
